Lesson 3
========================================================

***

### What to Do First?
Notes:

***

### Pseudo-Facebook User Data
Notes:

```{r Pseudo-Facebook User Data}
getwd()
list.files()
pf <- read.csv('pseudo_facebook.tsv', sep = '\t')
# read.delim('pseudo_facebook.tsv') unction defaults to the tab character as the separator between values and the period as the decimal character.

```

***

### Histogram of Users' Birthdays
Notes:

```{r Histogram of Users\' Birthdays}
#install.packages('ggplot2', dependencies = TRUE)
library(ggplot2)
```
```{r}
names(pf) #column names
qplot(x = dob_day, data = pf) +
  scale_x_continuous(breaks=1:31)
```
The ggthemes package was developed by Jeffery Arnold. Check out examples of the themes on the github page. 
Chris is using theme_minimal()with the font size set to 24, which is why his output might look slightly different than yours. You can set the same theme in R by running the following code, or you can set the theme to a choice of your own. 
theme_set(theme_minimal(24)) 

```{r}
#Run the following code in R to get other themes. 
#install.packages('ggthemes', dependencies = TRUE) 
library(ggthemes) 
```


### Faceting
Notes:
 Faceting in ggplot2  (http://www.cookbook-r.com/Graphs/Facets_(ggplot2)/)

```{r Faceting}
qplot(x = dob_day, data = pf) +
  scale_x_continuous(breaks = 1:31) +
  facet_wrap(~dob_month, ncol = 3)
# facet_wrap(~variable_to_facet_over)
# facet_grid(var_to_split_in_vertical_direction ~ var_to_split_in_horizontal) # more useful when you are faceting over two or more variables
```


### Be Skeptical - Outliers and Anomalies
Notes:
Outliers or anomalies could represent bad data or limitations of the data, but they could be accurate data about an extreme case.

***

### Moira's Outlier
Notes:
#### Which case do you think applies to Moira’s outlier?
Response:
Bad data of an extreme case
***

### Friend Count
Notes:

#### What code would you enter to create a histogram of friend counts?

```{r Friend Count}
qplot(x = friend_count, data = pf)
#idem
qplot(x = friend_count, data = pf, binwidth = 25) +
  scale_x_continuous(limits = c(0, 1000), breaks = seq(0, 1000, 100))
```

#### How is this plot similar to Moira's first plot?
Response:
Long tail plot
***

### Limiting the Axes
Notes:
Scales in ggplot2 
http://ggplot2.tidyverse.org/reference/scale_continuous.html

```{r Limiting the Axes}
qplot(x = friend_count, data = pf, xlim = c(0, 1000))
#idem
qplot(x = friend_count, data = pf, binwidth = 25) +
  scale_x_continuous(limits = c(0, 1000))
```

### Exploring with Bin Width
Notes:

***

### Adjusting the Bin Width
Notes:

### Faceting Friend Count
```{r Faceting Friend Count}
# What code would you add to create a facet the histogram by gender?
# Add it to the code below.
qplot(x = friend_count, data = pf, binwidth = 50) +
  scale_x_continuous(limits = c(0, 1000),
                     breaks = seq(0, 1000, 50)) +
  facet_wrap(~gender)
```

***

### Omitting NA Values
Notes:

```{r Omitting NA Values}
qplot(x = friend_count, data = subset(pf, !is.na(gender)), binwidth = 100) +
  scale_x_continuous(limits = c(0, 1000),
                     breaks = seq(0, 1000, 50)) +
  facet_wrap(~gender)
```

***

### Statistics 'by' Gender
Notes:

```{r Statistics \'by\' Gender}
table(pf$gender)
by(pf$friend_count, pf$gender, summary)
#by(variable, categorical variable, function)
```

#### Who on average has more friends: men or women?
Response:
women
#### What's the difference between the median friend count for women and men?
Response:
22

#### Why would the median be a better measure than the mean?
Response:
Median is more robust than the mean.
***

### Tenure
Notes:

```{r Tenure} 
# antiguedad
qplot(x = tenure, data = pf, binwidth = 30,
      color = I('black'), fill = I('#099DD9'))

# see ggplot theme documentation for color and fill
```

***

#### How would you create a histogram of tenure by year?

```{r Tenure Histogram by Year}
# tenure is in days
# similar to qplot(x = tenure, data = pf, binwidth = 365... but does not work well with breaks and limits...
# use .25 o 1
qplot(x = tenure/365, data = pf, binwidth = .25,
      color = I('black'), fill = I('#099DD9')) +
  scale_x_continuous(breaks = seq(1, 7, 1), limits = c(0, 7))
``` 

***

### Labeling Plots
Notes:

```{r Labeling Plots}
qplot(x = tenure/365, data = pf, binwidth = .25,
      xlab = "Number of years using facebook",
      ylab = "Numer of users in sample",
      color = I('black'), fill = I('#099DD9')) +
  scale_x_continuous(breaks = seq(1, 7, 1), limits = c(0, 7))
```

***

### User Ages
Notes:

```{r User Ages}
summary(pf$age)
qplot(x = age, data = pf, binwidth = .25,
      #xlab = "Number of years using facebook",
      #ylab = "Numer of users in sample",
      color = I('black'), fill = I('#099DD9')) + 
  scale_x_continuous(breaks = seq(13, 113, 10), limits = c(13, 113))
```

#### What do you notice?
Response:

***

### The Spread of Memes
Notes:

***

### Lada's Money Bag Meme
Notes:

***

### Transforming Data
When we have tails like this we might need to transform data in order to have a better comprehension of what is happening in the data (see patterns more clearly), like using logarithm or squared scales. A lot of common statistical techniques like linear regression are based on the assumption that variables have normal distributions, so by taking the log of a variable we can transform our data to turn it into maybe a normal distribution or something close to a normal distribution if we'll be using a linear regression.
***
```{r}
qplot(x = friend_count, data = pf)
summary(pf$friend_count)
summary(log10(pf$friend_count + 1)) # +1 to avoid zero, when calculating the mean, result negative infinitive
# another transformation
summary(sqrt(pf$friend_count))

#install.packages('gridExtra')
library('gridExtra')
p1 <- qplot(x = friend_count, data = pf)
p2 <- qplot(x = log10(friend_count + 1), data = pf)
p3 <- qplot(x = sqrt(friend_count), data = pf)
grid.arrange(p1, p2, p3, ncol=1)
```

### Transforming data alternative solution with ggplot
### Difference with ggplot: x and y variables inside aes (aesthetic) and adding geometry
Notes:

```{r}
p1 <- ggplot(aes(x = friend_count), data = pf) + geom_histogram()
p2 <- p1 + scale_x_log10()
p3 <- p1 + scale_x_sqrt()

grid.arrange(p1, p2, p3, ncol = 1)
```



### Add a Scaling Layer
Notes:

```{r Add a Scaling Layer}
logScale <- qplot(x = log10(friend_count), data = pf)
countScale <- ggplot(aes(x = friend_count), data = pf) + geom_histogram() + scale_x_log10()
grid.arrange(logScale, countScale, ncol = 2)
# Adding a scale to aplot too, difference is in the x scale labeling
qplot(x = friend_count, data = pf) +
  scale_x_log10()

```

***


### Frequency Polygons
Another type of plot to compare distributions, similar to histograms but they draw a curve to connect the counts of the histogram:

```{r Frequency Polygons}
qplot(x = friend_count, data = subset(pf, !is.na(gender)), 
      binwidth = 10, geom = 'freqpoly', color = gender) +
  scale_x_continuous(limits = c(0, 1000),
                     breaks = seq(0, 1000, 50)) 

# but who has more friends on average, we need proportions

qplot(x = friend_count, y = ..count../sum(..count..),
      data = subset(pf, !is.na(gender)), 
      xlab = 'Friend count',
      ylab = 'proportion of users with that friend count',
      binwidth = 10, geom = 'freqpoly', color = gender) +
  scale_x_continuous(limits = c(0, 1000),
                     breaks = seq(0, 1000, 50)) 

# equivalent
ggplot(aes(x = friend_count, y = ..count../sum(..count..)),
       data = subset(pf, !is.na(gender))) +
  geom_freqpoly(aes(color = gender), binwidth=10) +
  scale_x_continuous(limits = c(0, 1000), breaks = seq(0, 1000, 50)) +
  xlab('Friend Count') +
  ylab('Proportion of users with that friend count')

#Note that sum(..count..) will sum across color, so the proportions displayed are based on total users. To plot these proportions within each group, you can try y = ..density...
```

***

### Likes on the Web
Notes:

```{r Likes on the Web}
qplot(x = www_likes, data = subset(pf, !is.na(gender)), 
       geom = 'freqpoly', color = gender) +
  scale_x_log10()

by(pf$www_likes, pf$gender, sum)

```


***
### Histograms review

```{r Histograms review}
qplot(x = friend_count, data = subset(pf, !is.na(gender)), binwidth = 25)+
  scale_x_continuous(limits = c(0, 1000), breaks = seq(0, 1000,50))+
  facet_wrap(~gender)
```

### Box Plots
Notes:

```{r Box Plots}
qplot(x = gender, y = friend_count, 
      data = subset(pf, !is.na(gender)), 
      geom = 'boxplot') 

```

#### Adjust the code to focus on users who have friend counts between 0 and 1000.

```{r}
qplot(x = gender, y = friend_count, 
      data = subset(pf, !is.na(gender)), 
      geom = 'boxplot', ylim = c(0, 1000)) 
# alternative
qplot(x = gender, y = friend_count, 
      data = subset(pf, !is.na(gender)), 
      geom = 'boxplot') +
  scale_y_continuous(limits = c(0, 1000))
#better way
qplot(x = gender, y = friend_count, 
      data = subset(pf, !is.na(gender)), 
      geom = 'boxplot') +
  coord_cartesian(ylim = c(0, 1000))
```

***

### Box Plots, Quartiles, and Friendships
Notes:

```{r Box Plots, Quartiles, and Friendships}
qplot(x = gender, y = friend_count, 
      data = subset(pf, !is.na(gender)), 
      geom = 'boxplot') +
  coord_cartesian(ylim = c(0, 250))
# coord_cartesian for the graph to match the data from below table
by(pf$friend_count, pf$gender, summary)
```

#### On average, who initiated more friendships in our sample: men or women?
Response:
#### Write about some ways that you can verify your answer.
Response:
```{r Friend Requests by Gender}
qplot(x = gender, y = friendships_initiated, 
      data = subset(pf, !is.na(gender)),
      geom = 'boxplot') +
  coord_cartesian(ylim = c(0, 150))

by(pf$friendships_initiated, pf$gender, summary)
```


### Getting Logical
Notes:
Convert to binary variables from variables with a lot of zeroes

```{r Getting Logical}
summary(pf$mobile_likes)
summary(pf$mobile_likes > 0)

mobile_check_in <- NA
pf$mobile_check_in <- ifelse(pf$mobile_likes > 0, 1, 0)
pf$mobile_check_in <- factor(pf$mobile_check_in)

summary(pf$mobile_check_in)
        
```

### Percentage of user using mobile check in

```{r}
summary(pf$mobile_check_in)
sum(pf$mobile_check_in == 1) / length(pf$mobile_check_in)
        
```

### Analyzing One Variable
Reflection:
***
Analyzing one variable distribution to understand them with histograms, box plots and frequency plots.
Adjusting limits, axis transformation (logarithmical, squared, to obtain sometimes normal distributions for being able to make linear regressions)

Click **KnitHTML** to see all of your hard work and to have an html
page of this lesson, your answers, and your notes!