Lesson 6
========================================================

### Welcome
Notes:
See blog post for this rmd explanation:
https://solomonmessing.wordpress.com/2014/01/19/visualization-series-the-scatterplot-or-how-to-use-data-so-you-dont-get-ripped-off/
***

```{r}
library('ggplot2')
data(diamonds)
str(diamonds)
```


### Scatterplot Review

```{r Scatterplot Review}
qplot(x = carat, y = price, data = diamonds,
    xlim = c(0, quantile(diamonds$carat,0.99)),
    ylim = c(0, quantile(diamonds$price, 0.99)))+
  geom_point(fill = I('#F79420'), color = I('black'), shape = 21)
```

```{r Scatterplot Review ggplot syntax}
ggplot(aes(x = carat, y = price), data = diamonds) +
  geom_point(fill = I('#F79420'), color = I('black'), shape = 21) +
  #linear trim
  stat_smooth(method = 'lm') +
  #using 99 percentile
  xlim(0, quantile(diamonds$carat,0.99)) +
  ylim(0, quantile(diamonds$price, 0.99))
  
```


***

### Price and Carat Relationship
Response:
Non linear relationship, maybe exponential, maybe something else. Price increases with numbers of carats. If we paint a linear trend line, we can see that it doesn´t go through the center of the data in some key places.
***

### Frances Gerety
Notes:

#### A diamonds is


***

### The Rise of Diamonds
Notes:

***

### ggpairs Function
Notes:
Ggpairs provide:
-lower triangle grouped histograms for qualitative-qualitative pairs (using the y as grouping factor) and scatter plots for quantitative-quantitative pairs
-upper triangle grouped histograms for qualitative-qualitative pairs (using the x as grouping factor), boxplots for qualitative-quantitative pairs and correlation for quantitative-quantitative pairs.
```{r ggpairs Function}
# install these if necessary
#install.packages('GGally')
#install.packages('scales')
#install.packages('memisc')
#install.packages('lattice')
#install.packages('MASS')
#install.packages('car')
#install.packages('reshape')
#install.packages('plyr')

# load the ggplot graphics package and the others
library(ggplot2)
library(GGally)
library(scales)
library(memisc)

# sample 10,000 diamonds from the data set otherwise it will take too long
set.seed(20022012)
diamond_samp <- diamonds[sample(1:length(diamonds$price), 10000), ]
#ggpairs(diamond_samp, params = c(shape = I('.'), outlier.shape = I('.')))
ggpairs(diamond_samp,
  # to make the points smaller and clearer
  lower = list(continuous = wrap("points", shape = I('.'))),
  upper = list(combo = wrap("box", outlier.shape = I('.'))))
  #axisLabels = 'internal')
```

What are some things you notice in the ggpairs output?
Response:


***

### The Demand of Diamonds
Notes:

```{r The Demand of Diamonds}
plot1 <- qplot(data = diamonds, x = price, binwidth = 100, fill = I('#099DD9')) +
  ggtitle('price')
plot2 <- qplot(data = diamonds, x = price, binwidth = 0.01, fill = I('#F79420')) +
  ggtitle('price (log10)') + 
  scale_x_log10()
plot3 <- qplot(data = diamonds, x = log10(price), binwidth = 0.01, fill = I('#F79420')) +
  ggtitle('log10(price)') 
  

library(gridExtra)
library(grid)
grid.arrange(plot1, plot2, plot3, ncol = 2)
```

***

### Connecting Demand and Price Distributions
Notes:

***

### Scatterplot Transformation
On the log scale prices look less dispersed at the high end of carat size and price.
```{r Scatterplot Transformation}
qplot(carat, price, data = diamonds)+
  scale_y_continuous(trans = log10_trans())+
  ggtitle('Price (log10) by Carat')
```


### Create a new function to transform the carat variable
We can do better (we are trying to get a linear representation)
```{r cuberoot transformation}
# this is an R function. We need the inverse conversion too.
cuberoot_trans = function() trans_new('cuberoot', transform = function(x) x^(1/3),
                                      inverse = function(x) x^3)
```

#### Use the cuberoot_trans function
```{r Use cuberoot_trans}
ggplot(aes(carat, price), data = diamonds) + 
  geom_point() + 
  scale_x_continuous(trans = cuberoot_trans(), limits = c(0.2, 3),
                     breaks = c(0.2, 0.5, 1, 2, 3)) + 
  scale_y_continuous(trans = log10_trans(), limits = c(350, 15000),
                     breaks = c(350, 1000, 5000, 10000, 15000)) +
  ggtitle('Price (log10) by Cube-Root of Carat')
```

***

### Overplotting Revisited
Due mainly to rounding we have multiple points overlapped which obscures some of the density and sparsity of our data, the key points.

```{r Sort and Head Tables}
head(sort(table(diamonds$carat), decreasing = T))
head(sort(table(diamonds$price), decreasing = T))
```
You can make your points smaller, jittering them and adding some transparency.


```{r Overplotting Revisited}
# Add a layer to adjust the features of the
# scatterplot. Set the transparency to one half,
# the size to three-fourths, and jitter the points.

# If you need hints, see the Instructor Notes.
# There are three hints so scroll down slowly if
# you don’t want all the hints at once.

ggplot(aes(carat, price), data = diamonds) + 
  geom_point(alpha = 0.5, size = 0.75, position = 'jitter') + 
  scale_x_continuous(trans = cuberoot_trans(), limits = c(0.2, 3),
                     breaks = c(0.2, 0.5, 1, 2, 3)) + 
  scale_y_continuous(trans = log10_trans(), limits = c(350, 15000),
                     breaks = c(350, 1000, 5000, 10000, 15000)) +
  ggtitle('Price (log10) by Cube-Root of Carat')
```

***

### Other Qualitative Factors
Notes:

***

### Price vs. Carat and Clarity

Alter the code below.
```{r Price vs. Carat and Clarity}
# Adjust the code below to color the points by clarity.

# A layer called scale_color_brewer() has 
# been added to adjust the legend and
# provide custom colors.

# See if you can figure out what it does.
# Links to resources are in the Instructor Notes.

# You will need to install the package RColorBrewer
# in R to get the same colors and color palettes.

# install and load the RColorBrewer package
#install.packages('RColorBrewer')
library(RColorBrewer)

ggplot(aes(x = carat, y = price, color = clarity), data = diamonds) + 
  geom_point(alpha = 0.5, size = 1, position = 'jitter') +
  scale_color_brewer(type = 'div',
    guide = guide_legend(title = 'Clarity', reverse = T,
    override.aes = list(alpha = 1, size = 2))) +  
  scale_x_continuous(trans = cuberoot_trans(), limits = c(0.2, 3),
    breaks = c(0.2, 0.5, 1, 2, 3)) + 
  scale_y_continuous(trans = log10_trans(), limits = c(350, 15000),
    breaks = c(350, 1000, 5000, 10000, 15000)) +
  ggtitle('Price (log10) by Cube-Root of Carat and Clarity')
```

***

### Clarity and Price
Response: For a fixed number of carats we can see that diamonds with lower clarity al almost always cheaper than diamonds with higher clarity.

***

### Price vs. Carat and Cut

Alter the code below.
```{r Price vs. Carat and Cut}
ggplot(aes(x = carat, y = price, color = cut), data = diamonds) + 
  geom_point(alpha = 0.5, size = 1, position = 'jitter') +
  scale_color_brewer(type = 'div',
                     guide = guide_legend(title = 'Cut', reverse = T,
                                          override.aes = list(alpha = 1, size = 2))) +  
  scale_x_continuous(trans = cuberoot_trans(), limits = c(0.2, 3),
                     breaks = c(0.2, 0.5, 1, 2, 3)) + 
  scale_y_continuous(trans = log10_trans(), limits = c(350, 15000),
                     breaks = c(350, 1000, 5000, 10000, 15000)) +
  ggtitle('Price (log10) by Cube-Root of Carat and Cut')
```

***

### Cut and Price
Response: Most of the data is from diamonds with ideal cut, but we do see that most of the fair cuts are in the lower prices for a given carat.

***

### Price vs. Carat and Color

Alter the code below.
```{r Price vs. Carat and Color}
ggplot(aes(x = carat, y = price, color = color), data = diamonds) + 
  geom_point(alpha = 0.5, size = 1, position = 'jitter') +
  scale_color_brewer(type = 'div',
                     guide = guide_legend(title = 'Color',
                                          override.aes = list(alpha = 1, size = 2))) +  
  scale_x_continuous(trans = cuberoot_trans(), limits = c(0.2, 3),
                     breaks = c(0.2, 0.5, 1, 2, 3)) + 
  scale_y_continuous(trans = log10_trans(), limits = c(350, 15000),
                     breaks = c(350, 1000, 5000, 10000, 15000)) +
  ggtitle('Price (log10) by Cube-Root of Carat and Color')
```

***

### Color and Price
Response:Color does seem to explain some variance in the price.

***

### Linear Models in R
Notes: In R we can create models using the lm() function. x
lm(x~y) with x the outcome variable and y the explanatory variable.
Response:
In our case 
x = log(price)
y = carat ^ 1/3   (cube root)
***

### Building the Linear Model
Notes:
The I wrapper around the variables stands for as-is. In this case it tells R to use the expression inside the I function to tranform a variable before using it in the regression. This is instead of instructing R to interpret these symbols as part of the formula to construct the design matrix for the regression.
```{r Building the Linear Model}
m1 <- lm(I(log(price)) ~ I(carat^(1/3)), data = diamonds)
# updating the model
m2 <- update(m1, ~ . + carat)
m3 <- update(m2, ~ . + cut)
m4 <- update(m3, ~ . + color)
m5 <- update(m4, ~ . + clarity)
# mtable produces a table of estimates for several models.
mtable(m1, m2, m3, m4, m5, sdigits = 3)
```

Notice how adding cut to our model does not help explain much of the variance
in the price of diamonds. This fits with out exploration earlier.

***

### Model Problems
Video Notes:
OUR MODEL IS  (see column 5, the last model in the previous result)

log(price) = 0.415 + 9.144 * carat^(1/3) - 1.093 * carat....

Research:
(Take some time to come up with 2-4 problems for the model)
(You should 10-20 min on this)

Response:

***

### A Bigger, Better Data Set
Notes:

```{r A Bigger, Better Data Set}
#install.packages('bitops')
#install.packages('RCurl')
library('RCurl')
##library('bitops')

#diamondsurl = getBinaryURL("https://raw.github.com/solomonm/diamonds-data/master/BigDiamonds.Rda")
#load(rawConnection(diamondsurl))
load("BigDiamonds.rda")
```

The code used to obtain the data is available here:
https://github.com/solomonm/diamonds-data

## Building a Model Using the Big Diamonds Data Set
Notes:

```{r Building a Model Using the Big Diamonds Data Set}
# Your task is to build five linear models like Solomon
# did for the diamonds data set only this
# time you'll use a sample of diamonds from the
# diamondsbig data set.

# Be sure to make use of the same variables
# (logprice, carat, etc.) and model
# names (m1, m2, m3, m4, m5).

# To get the diamondsbig data into RStudio
# on your machine, copy, paste, and run the
# code in the Instructor Notes. There's
# 598,024 diamonds in this data set!

# Since the data set is so large,
# you are going to use a sample of the
# data set to compute the models. You can use
# the entire data set on your machine which
# will produce slightly different coefficients
# and statistics for the models.

# This exercise WILL BE automatically graded.

# You can leave off the code to load in the data.
# We've sampled the data for you.
# You also don't need code to create the table output of the models.
# We'll do that for you and check your model summaries (R^2 values, AIC, etc.)

# Your task is to write the code to create the models.
#
# DO NOT ALTER THE CODE BELOW THIS LINE (Reads in a sample of the diamondsbig data set)
#===========================================================================================
# diamondsBigSample <- read.csv('diamondsBigSample.csv')


# ENTER YOUR CODE BELOW THIS LINE. (Create the five models)
#===========================================================================================
diamondsbig$logprice = log(diamondsbig$price)
m1 <- lm(logprice ~ I(carat^(1/3)),
         data=diamondsbig[diamondsbig$price < 10000 &
                            diamondsbig$cert == "GIA",])
m2 <- update(m1, ~ . + carat)
m3 <- update(m2, ~ . + cut)
m4 <- update(m3, ~ . + color)
m5 <- update(m4, ~ . + clarity)
mtable(m1, m2, m3, m4, m5)

# DO NOT ALTER THE CODE BELOW THIS LINE (Tables your models and pulls out the statistics)
#===========================================================================================
suppressMessages(library(lattice))
suppressMessages(library(MASS))
suppressMessages(library(memisc))
models <- mtable(m1, m2, m3, m4, m5)
```


***

## Predictions

Example Diamond from BlueNile:
Round 1.00 Very Good I VS1 $5,601

```{r}
#Be sure you’ve loaded the library memisc and have m5 saved as an object in your workspace.
thisDiamond = data.frame(carat = 1.00, cut = "V.Good",
                         color = "I", clarity="VS1")
modelEstimate = predict(m5, newdata = thisDiamond,
                        interval="prediction", level = .95)
# We need to exponentiate the result of the model wince we took the log of price
exp(modelEstimate)
```

Evaluate how well the model predicts the BlueNile diamond's price. Think about the fitted point estimate as well as the 95% CI.


***

## Final Thoughts
Notes:

***

Click **KnitHTML** to see all of your hard work and to have an html
page of this lesson, your answers, and your notes!

